In this chapter, we discuss what we observe once the simulations are running in practice, and what the measurements tell us. We first focus on the burn-in phase, since it sets the point from which our data can be treated as stationary. We then turn to the scaling exponents extracted from avalanche statistics and comment on how robust they are across models and system sizes. Finally, we summarise the main trends in the results and highlight the most likely sources of systematic error.
\subsection{Burn-in dynamics and monitoring}

To measure avalanche statistics, we first need the system to settle into a steady state. The early part of the run is the burn-in phase. In this phase, average quantities still change with time. We monitor this by tracking the mean energy (mean slope) $\langle z\rangle(\tau)$ and waiting until it stops showing a clear trend.

Models A and B behave differently because they introduce energy differently. Model A uses nonconservative driving, so it increases the total energy directly and the system reaches the stationary energy density relatively quickly. Model B uses a conservative perturbation. In the bulk it keeps the total energy roughly constant and mainly redistributes what is already there. Starting from an empty lattice, this makes model B take much longer to build up to the stationary level than model A.

A reasonable explanation is that, with our boundary conditions, conservative perturbations only create a net energy increase through boundary effects. The conservative update subtracts energy from shifted neighbour sites. At the boundaries some of these sites are clamped (especially at the lower boundary), so this subtraction can be partly blocked. This means the perturbation can effectively add energy only when it interacts with the edges, while perturbations in the bulk are close to energy-neutral. If most of the net input comes from the edges, then the build-up from $z_0=0$ is naturally slower and more sensitive to how the boundaries are implemented.

\subsection{Reducing burn-in by changing the initial condition}

The long burn-in in model B is mainly due to how slowly the system fills up when starting from $z_0=0$. A direct way to shorten it is to start closer to the stationary energy density. This motivates the alternative initial conditions: random initialization $z_0\sim U(0,z_c)$ and maximally stable initialization $z_0=z_c$. Both preload the lattice with energy, so fewer driving steps are needed before $\langle z\rangle(\tau)$ stops drifting.

These initial conditions also reduce the time spent in a clearly nonstationary regime, where avalanches are not representative of the steady state. The benefit is largest for conservative driving, where the fill-up stage is slowest. For cases that equilibrate quickly, the burn-in time still cannot drop below the built-in minimum $T_{\min}$ set by the stationarity check procedure.

\subsection{Scaling exponent measurements and comparison to literature}
After preparing each system in Table~\ref{tab:model_configurations} in the
critical state, we measure the avalanche observables $t$, $s$ and $l$.
From these measurements we estimate the PDFs and fit power laws in the
scaling region (see the example fits in Fig.~\ref{fig:pdf-plot-example}).

On the double-logarithmic PDF plots, the scaling region appears as an
approximately straight line. At small $t,s,l$ we do not expect perfect
scaling because of discretization effects and because avalanches are still
``small''. At large $t,s,l$ we see a cutoff due to finite system size and
limited statistics (especially for the $d=4$ systems in
Table~\ref{tab:model_configurations}, where we used fewer measurements).
Because of these effects, the choice of the fitting window has a direct
impact on the fitted slope (as visible in Fig.~\ref{fig:pdf-plot-example}).


The fitted PDF exponents for the open nonconservative systems (configs
A, D, G) are listed in Table~\ref{tab:scaling_exp_open_nonconservative}.
The corresponding results for the open conservative systems (configs
B, E, H) are given in Table~\ref{tab:scaling_exp_open_conservative}, and for
the closed nonconservative systems (configs C, F, I) in
Table~\ref{tab:scaling_exp_closed_nonconservative}.
Overall, the fitted values follow the expected trend with dimension: the
distributions become steeper with increasing $d$, and the scaling window is
typically shorter in higher dimensions, which increases the fit uncertainty.
This is most visible for the $l$-fits (the exponent $\lambda$), where the
cutoff sets in early and the statistics at large $l$ are weaker.

Comparing boundary conditions for the nonconservative perturbation (open:
Table~\ref{tab:scaling_exp_open_nonconservative} vs.\ closed:
Table~\ref{tab:scaling_exp_closed_nonconservative}), the exponents are
similar within the uncertainties. This suggests that changing open to closed
boundaries does not strongly change the bulk scaling, as long as avalanches
can still dissipate energy by the chosen dynamics.

Comparing perturbation schemes at open boundaries (nonconservative:
Table~\ref{tab:scaling_exp_open_nonconservative} vs.\ conservative:
Table~\ref{tab:scaling_exp_open_conservative}), we observe systematic shifts
in the fitted exponents. In particular, the conservative perturbation tends
to give steeper PDFs, which means very large avalanches are relatively less
likely in that setup.


The conditional expectation value fits are illustrated in
Fig.~\ref{fig:cond-plot-example}. The measured conditional exponents for the
open nonconservative systems are summarized in
Table~\ref{tab:scaling_exp_open_nonconservative_gamma}, for open
conservative in Table~\ref{tab:scaling_exp_open_conservative_gamma}, and for
closed nonconservative in Table~\ref{tab:scaling_exp_closed_nonconservative_gamma}.
In general, the relations involving $s$ and $t$ show the most stable scaling
regions, while relations that involve $l$ tend to have larger uncertainties.
A simple reason is that large-$l$ events are rare, so conditional averages
at large conditioning values have larger noise, and the fitted slope depends
more on the exact window choice (as also visible in
Fig.~\ref{fig:cond-plot-example}).


The reference values reported by Christensen et al.\ can be compared
directly to our fitted exponents in
Tables~\ref{tab:scaling_exp_open_nonconservative}--\ref{tab:scaling_exp_closed_nonconservative_gamma}.
Overall, our results show the same qualitative trends with dimension and
with the choice of perturbation/boundary conditions. Remaining differences
are expected from finite-size effects, the limited scaling window, and the
fact that exponent estimates depend on the fitting procedure and on the
chosen range (see Fig.~\ref{fig:pdf-plot-example} and
Fig.~\ref{fig:cond-plot-example}).